\chapter{Deployment Models}\label{ch:resourcemodels}


\section{Introduction}

  This section describes language features related to deployment
  components.  Most of these features are implemented on the Maude
  backend and ignored on the Java backend.

  ABS models can be augmented with resource information and the effects
  of resources during execution can be simulated on the Maude backend.
  Resources are contained in \emph{deployment components}, which provide a
  context in which concurrent object groups and their objects execute.
  Deployment components can be used to model servers, virtual machines,
  nodes in a network, and basically any entity that constrains execution
  of code running in its context.

  At the moment we only model CPU capacity.  A deployment component (DC)
  can contain a certain ``CPU capacity'', which provides an abstract model
  of speed relative to other DCs; i.e., a cpu capacity of 50 is ``twice
  as fast'' as 25.

  Executing statements costs a certain number of resources, which are
  refilled in each time interval -- since CPU speed is related to
  performance and hence time, simulations involving performance
  characteristics must be done using the {{{./timedmodeling.html}timed
  interpreter}}.

\section{Language Elements}

  All identifiers described in the following are contained in the module
  \absinline{ABS.DC}.

\subsection{Resource Types}

   Deployment components carry a specific capacity, which is a possibly
   infinite number of CPU ``cycles''.  The partial function \absinline{capacity}
   returns the number of resources, if finite.

\begin{absexample} 
data DCData = InfCPU | CPU(Int capacity);
\end{absexample} 

\subsection{Deployment Components}

   A deployment component (``DC'') is modeled by an ABS class and
   interface \absinline{DeploymentComponent} which is contained in the
   standard library in the module \absinline{ABS.DC}.  DCs are initialized
   with resource data which is used by the Maude backend to simulate cpu
   usage and performance behavior of models.

\begin{absexample} 
interface DeploymentComponent {
    DCData available();
    Int load(Int periods);
    DCData total();
    Unit transfer(DeploymentComponent target, Int amount);
    Unit decrementResources(Int amount);
    Unit incrementResources(Int amount);
}

class DeploymentComponent (String description, DCData cpu)
implements DeploymentComponent {
    ...
}
\end{absexample} 

\begin{itemize}
\item The call \absinline{available()} returns the number of currently
 available resources.  (The value will change during execution,
     hence code using this method should be aware of the obvious race
     condition.)
\item The call \absinline{load(n)} returns the \emph{average load} for the
     last n time periods as an integer between 0 and 100.
\item The method \absinline{total()} returns the number of resources currently
     available per time unit.
\item The method \absinline{transfer(target, amount)} transfers a number of
     resources from the current DC to \absinline{target}.  If either the
     current or the target DC have an infinite number of resources,
     their respective values are not changed.
   \item The methods \absinline{decrementResources} and \absinline{incrementResources}
     are used to implement \absinline{transfer}, but can also used on their
     own to increment or decrement the amount of resources of the
     current deployment component.
\end{itemize}

\subsection{Using Deployment Components}

   An optional annotation \absinline{[DC: x]} attached to a \absinline{new cog}
   statement expresses that the new object's cog will run in the context
   of deployment component \absinline{x}.  By default, a new cog runs in the
   same context as the process that generates it.  The main block runs
   in a DC with no resource limits.

\begin{absexample} 
interface Server { ... }

class ServerImp implements Server { ... }

{
  DeploymentComponent machine = new DeploymentComponent("remote",
      CPU(20));
  [DC: machine] Server x = new cog ServerImp();  // running in dc `machine'
  Server y = new cog ServerImp();                // running in current DC
}
\end{absexample} 

\subsection{Accessing the Current Deployment Component}

\begin{absexample} 
def DeploymentComponent thisDC() = builtin;
\end{absexample} 

   The \absinline{thisDC()} function returns the current deployment
   component, i.e., the one that was given as annotation at the \absinline{new
   cog} object instantiation.  \absinline{thisDC} returns \absinline{null} if no
   deployment component was specified.  (Due to implementation issues,
   models containing \absinline{thisDC()} cannot be currently compiled on the
   Java platform.)

\subsection{Expressing statement costs}

   The annotation \absinline{[Cost: x] s} expresses that executing the
   statement \absinline{s} will consume \absinline{x} CPU resources.  If the current
   DC does not contain enough resources, the process will \emph{block} until
   the next time unit.

\subsection{Default statement costs}

   By default, statement execution costs nothing (i.e., the machine is
   infinitely fast), which preserves the normal semantics of timed ABS
   where only explicit \absinline{duration} or \absinline{await duration} statements
   advance the clock.  By compiling ABS models with the command-line
   parameter \absinline{-defaultcost=n}, all statements without an explicit
   \absinline{[Cost: x]} annotation will take \absinline{n} resources to execute.
   This approximates ``normal'' CPU behavior a bit more closely, but still
   disregards the cost of evaluating expressions (assignment of a
   complex expression will take more time on a real machine than
   assignment of a constant).

\subsection{Changing the deployment component}

   A statement \absinline{movecogto dc;} transfers the current cog and all its
   objects to the deployment component \absinline{dc}.  On backends without
   resource modeling features (Java, Scala) this translates to a
   \absinline{skip}.

\section{Future Work}

The \absinline{thisDC()} function is currently undefined in the Java
backend and will lead to a compile-time error.
