\chapter{Lexical Structure}
\label{ch:lexical structure}
This section describes the lexical structure of the ABS language.
ABS programs are written in Unicode.\footnote{\url{http://www.unicode.org}}

\section{Line Terminators and White Spaces}
Line terminators and white spaces are defined as in Java.
%\begin{absgrammar}
%LineTerminator ::= \n | \r | \r\n
%WhiteSpace     ::= LineTerminator | ' ' | \t | \f
%\end{absgrammar}
%
\begin{abssyntax}
\NT{LineTerminator}
\concrDefn{} \verb_\_\TR{n}  ~|~ \verb_\_\TR{r} ~|~ \TR{rn} 
\\
\NT{WhiteSpace} \concrDefn{}
  \NT{LineTerminator} ~|~ \verb*_ _ ~|~ \verb_\_\TR{t} ~|~ \verb_\_\TR{f}
\end{abssyntax}

\section{Comments}
Comments are code fragments that are completely ignored and have no semantics in the ABS language.
ABS supports two styles of comments:
\emph{end-of-line comments} and \emph{traditional comments}.

\subsection{End-Of-Line Comments}
An end-of-line comment is a code fragment that starts with two slashes, e.g., \verb_// text_. All text that follows \verb_//_ until the end of the line is treated as a comment.

\begin{absexample}
// this is a comment
module A; // this is also a comment
\end{absexample}

\subsection{Traditional Comments}
A traditional comment is a code fragment that is enclosed in \absinline{/* */}, e.g., \absinline{/* this is a comment */}.
Nested traditional comments are not possible.

\begin{absexample}
/* this
is a multiline
comment */  
\end{absexample}


\section{Identifiers}
ABS distinguishes \emph{identifier} and \emph{type identifier}.
They differ in the first character, which must be a lower-case character
for identifiers and an upper-case character for type identifiers.
%
%\begin{absgrammar}
%Identifier     ::= [:lowercase:] {[:letter:] | [:digit:] | '_'}
%TypeId         ::= [:uppercase:] {[:letter:] | [:digit:] | '_'}
%\end{absgrammar}
%
\begin{abssyntax}
\NT{Identifier}  \defn
  \text{[:lower:]} \MANYG{\text{[:alpha:] ~|~ [:digit:] ~|~ \TR{\_}}}
\\
\NT{TypeId}      \defn
  \text{[:upper:]} \MANYG{\text{[:alpha:] ~|~ [:digit:] ~|~ \TR{\_}}}
\end{abssyntax}

\section{Keywords}
The following words are keywords in the ABS language and are \emph{not}
regarded as identifiers.

\noindent
\begin{center}
\begin{tabular}{llllll}
\TR{adds} & \TR{after} & \TR{assert} & \TR{await} & \TR{builtin} & \TR{case} \\
\TR{core} & \TR{class} & \TR{data} & \TR{def} & \TR{delta} & \TR{else} \\
\TR{export} & \TR{features} & \TR{from} & \TR{get} & \TR{hasField} & \TR{hasInterface} \\
\TR{hasMethod} & \TR{if} & \TR{implements} & \TR{import} & \TR{in} & \TR{interface} \\
\TR{let} &  \TR{local} & \TR{modifies} & \TR{module} & \TR{new} & \TR{null} \\
\TR{product} & \TR{productline} & \TR{removes} & \TR{return} & \TR{skip} & \TR{suspend} \\
\TR{this} & \TR{type} & \TR{when} & \TR{while} & &\\ 
\end{tabular}
\end{center}

\section{Literals}
A \emph{literal} is a textual representation of a value.
ABS supports four kinds of literals, \emph{integer literals}, \emph{string literals}, the \emph{this literal}, and the \emph{null literal}.

\begin{abssyntax}
\NT{Literal}       \defn \NT{IntLiteral}
                ~|~ \NT{StringLiteral}
                ~|~ \NT{ThisLiteral}
                ~|~ \NT{NullLiteral}
                \\
\NT{IntLiteral}    \defn 0 ~|~ [1-9]\MANY{[0-9]}\\
\NT{StringLiteral} \defn \TR{"} \MANY{\NT{StringCharacter}} \TR{"}\\
\NT{ThisLiteral}   \defn \TR{this}\\
\NT{NullLiteral}   \defn \TR{null}
\end{abssyntax}

\noindent
Where a \NT{StringCharacter} is defined as in the Java language \cite[p.~28]{gosling96}

\section{Separators}
The following characters are \emph{separators}:
\begin{verbatim}
    (   )   {   }   [   ]   ,   ;   :
\end{verbatim}

\section{Operators}
The following tokens are \emph{operators}:
\begin{verbatim}
    ||   &&   ==   !=   <   >   <=   >=   +   -   *   /   %   !   &   
\end{verbatim}
% .   !   =   _
